Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-08 forecastflows.tex
modify box, reduce common workflow fig size
Some edits but seems good to get feedback on!

\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology } 
\date{\today}
\author{Victor Van der Meersch, J. Regetz, T. J. Davies$^*$ \& EM Wolkovich}
\maketitle
$^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author.

@@

Anthropogenic drivers are reshaping natural systems, with impacts expected to increase in coming decades \citep{Diaz2019}. Climate change will likely accelerate biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}. Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date alongside the skill to forecast future dynamics. % These requirements in turn have created a growing need for more global ecological data, and the types of models that can robustly handle such data.

Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets and another focused on forecasting from generally distinct datasets or mechanistic models based on less data. Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced with massive geographic, temporal and taxonomic biases. Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting. 
Instead forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning. The current outcome from these approaches is no clear agreement on current species trends, with ongoing debates on the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012}.

We argue that current debates and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, forecasting-focused process-based models often develop by adding a new separate layer, disconnected from the original research aim, the data stream and the previous scientific insights, because they fail to examine the model as a functioning whole and thus ignore major degeneracies.

Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data. Towards this aim, here we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

Quantitative science relies on a model-based framework, to confront hypothesis and data \citep{}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses. But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}.

This reality should drive researchers to use more robust and coherent methods, but the current workflows combined with the challenges ecologists are facing instead appear to be leading to persistent problems. 
Trend estimation has focused mostly on fitting the model to empirical data in a uni-directional way that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig. 1). For forecasting, researchers focus on making predictions with increasingly complex models, making the steps for model building and parameterization increasingly obscured (Fig. 1). Today, researchers often calibrate different parts of the model separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole. Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow. 
% mention the disconnection between trends and forecasts?

@@

We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation (Fig 1) could reduce many of these problems and thus improve ecological science.
The first step of the workflow is to define an explicit research question and formulate hypotheses (Fig 2). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step help us to think about the mechanisms that could generate the data we observe, including the observational error. Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and designed to answer our research question. The general idea is to start with a relatively simple model, that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.

With a model in place the next step focuses on testing and understanding it via data simulation. `Fake' or `test' data are generated directly from the model by fixing parameters to some values (which is straightforward if the parameters are interpretable) and from fake predictor data. This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model. For example, if a researcher realizes their empirical is geographically biased, that should be built into the model and thus then into this data simulation step. 
We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. 
Once we are confident about our model structure, we can incorporate real data. This way, we obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. This leads to the second data simulation step, this time using our fitted model parameters to generate predictions. This retrodictive check allows model output to be compared to observations. 
The workflow replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting. Any parameter (such as a trend estimate) must be carefully interpreted alongside others. %such as the  global mean or intercept and the estimate variance
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

\begin{figure}[h]
	\centering
    \hspace*{-1.5cm}
	\includegraphics{figures/figure_worflow}
	\caption{} 
	\label{fig:modeldata}
\end{figure}

@@

Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}. Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines, and has led to a debates within ecology about whether such analyses undermine public confidence in science  \citep{gonzalez2016estimating}. While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.

An improved workflow that required retrodictive checks and data simulation would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make what now appear as major discrepancies more obviously shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks (or whatever we call them)---are part of the process of science, and critical to testing for what may be missing in a model. We expect retrodictive checks on most published trend analyses would highlight major missing components in these models (expand here?? ADD example?), and drive changes both in the models themselves and in the simulated data to check the models. While ecologists have started to use simulated data more to understand potential limitations of their models and data combined, this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility
% Mention how low R2 are in ecology?
% Simulating data to verify and test models in trend analyses is currently rarely reported, which we believe makes it easier for ecologists to fit poor models and in turn debate them.. 
% Mention elephants rebounding...

@@

% vvdm07april: need a topic sentence
Increasing model complexity has not necessarily led to reduced uncertainty. In climate modeling, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

Applying the workflow to process-based models is a key for opening the black box. It would serve as a guide through the successive steps of model development. In particular, incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing strong degeneracies in the model design. This approach would force researchers to begin with a simpler version of the model, providing a clear pathway to support---or reject---the additional complexity and new parameters along the iterative development of the model. 
Model calibration would no longer be just a hidden aspect of model building but a step as crucial as forecasting to gain new ecological insights. It would help properly take into account any issues regarding non-identifiability. This could involve reformulating the mathematical structure of some processes, making new hypotheses to target the right level of complexity, or incorporating more expert knowledge to better constrain calibration. Making the model more tractable would also naturally facilitate the recognition and quantification of parameter uncertainty, as well as its propagation into model projections

Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification. This shift would present a significant challenge---as it would likely reveal many issues related to model degeneracies and data limitations before achieving robust inference. But ultimately, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.

@@

\centerline{\bf Common solution!}
\vspace{0.2cm}
\noindent\makebox[\textwidth][c]{
\begin{minipage}{0.8\linewidth}
    \includegraphics[width=\linewidth]{figures/iterativeworkflow_details}

    % \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
    \vspace*{1mm}
\end{minipage}}

@@

We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
One well-known hurdle is that the pressure to publish academic work, which can lower research standards and make the added effort of this workflow seem ill-placed. This may be especially true for those who see science rewarded mostly through the shear number of publications. But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help. Further, growing concerns about how reproducible science is, especially in ecology where samples sizes are low and effects likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, as more value is placed on research that carefully developed, openly shared (for data and code) and acknowledges its uncertainty towards the aim of improving future data collection and model development (see Fig).

\subsection{Comprehensive training}
% emw03april: massively shorten?
Advancing ecology to where most researchers see models built more flexibly from their understanding of their ecological systems will not happen rapidly without a major shift in training, however. Much of ecology still divides the world into training for those focused on several groups. First, those who will conduct field and lab studies often learn a suite of particular tests matched to particular experiment designs or to simply match their predictor and response data to a flow diagram of variable types (add EXAMPLE and reference, something like: categorical $y$ and binary $y$ means ... I dunno, chi square?) and are left adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is. Instead, they are expected to collaborate with others when they need more complex models, which is a second group of ecological researchers---those who focus on complex statistical models and were thus trained more in model development, but often for highly specific applications (e.g., wildlife population estimates) where they may rarely build totally new generative models. Separate from these groups, process-based modelers do their own thing. 
% process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
% hmm some people talk about simulation models (for process-based models), I agree but I'm afraid in would not be clear with the data simulation we try to highlight
Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data. While specialization is valuable, we argue the fundamental training in ecology has overly-siloed these groups and prevented more rapid progress.

Training all ecologists in our proposed workflow would break down barriers between different groups of ecologists today with likely major benefits for science. Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in our workflow would focus on learning to generate questions and then models, and then to simulate data from them. This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them. Through this and retrodictive checks most ecologists would more easily think through what parameters are most critical to their question and or aim (e.g., management) and also gain a much stronger connection to the level of uncertainty in many of ecological estimates. Empirical ecologists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models. Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting. 
% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches to better forecasts or something.

>>>>

\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology } 
\date{\today}
\author{Victor Van der Meersch, J. Regetz, T. J. Davies$^*$ \& EM Wolkovich}
\maketitle
$^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author (though Lizzie used his LPI example).

@@

Anthropogenic drivers are reshaping natural systems \citep{Diaz2019}. Impacts are projected to increase in coming decades, as climate change accelerates biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}. Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date alongside the skill to forecast future dynamics. % These requirements in turn have created a growing need for more global ecological data, and the types of models that can robustly handle such data.

Meeting these policy needs has led often to two separate paths: one focused on estimating trends from new global datasets and another focused on forecasting from generally distinct datasets or mechanistic models based on less data. Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced with massive geographic, temporal and taxonomic biases. Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting. 
Instead forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning. The current outcome from these approaches is no clear agreement on current species trends, with ongoing debates on the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012}.

We argue that current debates and diverging forecasts are driven in large part by the incoherent and disconnected workflows used today in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Research estimating biodiversity trends has become focused on methodological aspects because the current workflow fails to examine the gap between ideal and available data, and rarely tests for predictive accuracy that could scale up to allow forecasting. At the same time, forecasting-focused process-based models often develop by adding new separate layers or components. These new parts are often disconnected from the original research aim, its data stream and the previous scientific insights, because current approaches rarely examine the model as a functioning whole and thus ignore major degeneracies.

Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data. Towards this aim,  we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses. But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}.

This reality should drive researchers to use more robust and coherent methods, but the current workflows combined with the challenges ecologists are facing instead may lead to persistent problems. 
Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig. \ref{fig:modeldata}). For forecasting, researchers have focused on making predictions with increasingly complex models, increasingly obscuring the steps underlying model building and parameterization (Fig. \ref{fig:modeldata}). Today, researchers often calibrate different parts of the model separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole. Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow. 
% mention the disconnection between trends and forecasts?

@@

We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation (Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (Fig. \ref{fig:workflowy}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step should help focus on the mechanisms that could generate the data we observe, including the observational error. Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and is designed to answer our research question. The general idea is to start with a relatively simple model that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.

With a model in place the next step focuses on testing and understanding it via data simulation. `Fake' or `test' data are generated directly from the model by fixing parameters to some values (which is straightforward if the parameters are interpretable) and from fake predictor data (Fig. \ref{fig:workflowy}). This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model. For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.

Once we are confident about our model structure, we can incorporate real data. This way, we obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. This leads to the second data simulation step, this time using our fitted model parameters to generate predictions. This retrodictive check allows model output to be compared to observations. 
The workflow replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting. Any parameter (such as a trend estimate) must be carefully interpreted alongside others. %such as the  global mean or intercept and the estimate variance
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

\begin{figure}[h]
	\centering
    \hspace*{-1.5cm}
	\includegraphics{figures/figure_worflow}
	\caption{Here we could have a caption like ... Our workflow builds on others (with citations) to highlight the integrated nature of stuff in ecology. We can work on this once we get feedback from mighty co-authors. Also, should we list the steps on the figure? What do people think?} 
	\label{fig:workflowy}
\end{figure}

@@

Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}. While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced through use of an improved workflow. % Further, an improved workflow could give more rapid and coherent estimates, which could make it easier for  policy-makers to develop and advance initiatives aimed at slowing declines.
% Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines \citep{gonzalez2016estimating}. While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.

An improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make what now appear as major discrepancies more obviously shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model. We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models. While ecologists have started to use simulated data more to understand potential limitations of their models and data combined, this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility.
% expand on 'highlight major missing components in these models'? ADD example?
% Simulating data to verify and test models in trend analyses is currently rarely reported, which we believe makes it easier for ecologists to fit poor models and in turn debate them.. 
% Mention elephants rebounding...

@@

% vvdm07april: need a topic sentence
Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In climate modeling, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

Applying the workflow to process-based models would help open black box. Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing potentially large degeneracies in the model design. Uncovering these degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight degeneracies by requiring more explicit model calibration, which is currently hidden within opaque `model building,' making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated). Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.

Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.

This shift would present a significant challenge---as it would likely reveal many issues related to model degeneracies and data limitations before achieving robust inference. But ultimately, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support.

@@

\centerline{\bf A common workflow to bridge trend estimation and forecasting}
\vspace*{-3mm}
{\begin{multicols}{2}
\begin{minipage}[t]{\linewidth}
\includegraphics[width=\linewidth]{figures/iterativeworkflow_details}

\columnbreak
\vspace*{1mm}
Blabla \\
We say here briefly how great the new approach is or something.

@@

We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed. This may be especially true for those who see science rewarded mostly through the shear number of publications. But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help. Further, growing concerns about how reproducible science is, especially in ecology where samples sizes are low and effects likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, as more value is placed on research that carefully developed, openly shared (for data and code) and acknowledges its uncertainty towards the aim of improving future data collection and model development (see Fig).

\subsection{Comprehensive training}
Advancing ecology to where most researchers use models built more flexibly from ecological theory and insights applied to their ecological systems will not happen rapidly without a major shift in training. Much of ecology still divides the world into training for those focused on several groups: those who gather data and learn a limited set of pre-built models versus those who develop more complex models. In ecological training today may researchers who conduct field and lab studies often learn a highly limited set of particular tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test). Such training leaves a large number of ecologists adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is.

Instead, ecologists trained in a limited set of tests are expected to collaborate with others when they need more complex models, who were trained more in model development (though often for highly specific applications, such as wildlife population estimates, where they may rarely develop entirely new generative models). These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models. % process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data. While specialization is valuable, we argue the fundamental training in ecology has overly-siloed these groups and prevented more rapid progress.

Training all ecologists in our proposed workflow would break down barriers between these different groups with likely major benefits for science. Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in our workflow would focus on learning to generate questions and then models, and then to simulate data from them. This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them. Through this and retrodictive checks most ecologists would more easily think through what parameters are most critical to their question and or aim (e.g., management) and also gain a much stronger connection to the level of uncertainty in many of ecological estimates. Empirical ecologists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models. Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting. 
% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches to better forecasts or something.
